---
title: "Classification"
author: "Kostas Ntagiantas"
date: "July 18, 2018"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = TRUE, echo = FALSE, results = "hide", warning = 
                        FALSE, message = FALSE, cache.lazy = FALSE, fig.align = 
                        "center" , cache = FALSE)
library(knitr)
library(tidyverse)

code.name <- "Classification" 
data.dir <- file.path(getwd(), "data", code.name)
results.dir <- file.path(getwd(), "results", code.name)
experiment <- c("GSE7127", "GSE36133", "GSE22301", "GSE35388")

#Adding all useful things
#load(file.path(data.dir, "ClassificationInput.RData"))
load(file.path(data.dir, "finalData_Classification.RData"))

```

## Classification of healthy/cancerous cells using SVM

With the previous analysis the most abundant differentially expressed genes have been found and can now be used as features to classify normal from melanoma cell lines. The number of features depends on the threshold selection of the previous section.

```{r Features presented, results = 'markup'}
library(kableExtra)

data.frame( `Lax threshold` =paste( (genes.mel.healthy %>% filter(Set != "From all") %>% 
  group_by(Set, Gene) %>% summarise(n = n()) %>% 
    filter(n > threshold[1]) %>% 
    ungroup() %>% 
    group_by(Gene) %>% summarise(PresentIn = n() ) %>% 
    filter(PresentIn == 6) %>% .$Gene),collapse = ", "),
  
  `Medium threshold` = paste((genes.mel.healthy %>% filter(Set != "From all") %>% 
  group_by(Set, Gene) %>% summarise(n = n()) %>% 
    filter(n > threshold[2]) %>% 
    ungroup() %>% 
    group_by(Gene) %>% summarise(PresentIn = n() ) %>% 
    filter(PresentIn == 6) %>% .$Gene),collapse = ", "),
 
   `Strict threshold` = paste( (genes.mel.healthy %>% filter(Set != "From all") %>% 
  group_by(Set, Gene) %>% summarise(n = n()) %>% 
    filter(n > threshold[3]) %>% 
    ungroup() %>% 
    group_by(Gene) %>% summarise(PresentIn = n() ) %>% 
    filter(PresentIn == 6) %>% .$Gene),collapse = ", ")) %>% 
  kable(., align = "c", caption = "Genes used as features for classification") %>% kable_styling()

```


```{r SVM Melanoma VS Healthy, eval = FALSE, include = FALSE}
library(e1071)
library(caret)
library(mccr)
#threshold values for genes selection
threshold <- c(260, 540, 800)

#Cross validation and mcc function
cvsvm<- function(samples.index, df, nfolds){
  dfused = df[samples.index,]
  folds = createFolds(dfused$Status, k = nfolds)
  accuracy = 0
  trueControl = 0
  mcc = 0
  for(i in 1:length(folds)){
    test = dfused[folds[[i]],]
    train = dfused[-folds[[i]],]
    model_svm = svm(Status ~ . , train, type = "C")
    predict_svm = predict(model_svm, test[,-1])
    conf.matrix = table(predict_svm, test$Status)
    accuracy = accuracy + 100*(sum(diag(conf.matrix))/sum(conf.matrix))
    trueControl = trueControl + conf.matrix[1,1]
    mcc = mccr(as.integer(test$Status == "Control"), as.integer(predict_svm == "Control")) + mcc
  }
  value = c(accuracy, trueControl, mcc )/nfolds
  
  return(value)
} 

# df_main is the data frame that will be used in all downstream analysis
tmp <- pheno.global %>% 
  mutate(Status = gsub("Immortalized", "Control", Status)) %>% 
  column_to_rownames(var = "geo_accession") %>% 
  dplyr::select(Status)

df_main <- global.exp.corrected %>% column_to_rownames(var = "SYMBOL") %>% 
  t(.) %>% as.data.frame(.) %>% 
  cbind(tmp, .) 

control.index <- which(df_main$Status == "Control")
case.index <- which(df_main$Status == "Melanoma")

feature.genes <- genes.mel.healthy %>% filter(Set != "From all") %>% 
  group_by(Set, Gene) %>% summarise(n = n()) %>% 
    filter(n > threshold[1]) %>% 
    ungroup() %>% 
    group_by(Gene) %>% summarise(PresentIn = n() ) %>% 
    filter(PresentIn == 6) %>% .$Gene

#Testing accuracy with any number of melanoma samples
accuracy.list <- list()
for(i in 6:length(case.index)){
  set.seed(1)
  accuracy.list[[i-5]] <- replicate(10,{
  samples.index <- c( control.index, sample(case.index, i))
  cvsvm(samples.index = samples.index, df_main[,c("Status",feature.genes)], nfolds = 6)
  })
}
svm.results <- lapply(accuracy.list, rowMeans) %>% 
  purrr::reduce(.x = ., .f = rbind) %>% as.data.frame(.) %>% 
  dplyr::rename(Accuracy = V1, TrueControls = V2, MCC = V3) %>% 
  mutate(Features = "35-lax", MelSamples = c(6:length(case.index)))

#Applying a medium threshold to select the feature genes
feature.genes <- genes.mel.healthy %>% filter(Set != "From all") %>% 
  group_by(Set, Gene) %>% summarise(n = n()) %>% 
    filter(n > threshold[2]) %>% 
    ungroup() %>% 
    group_by(Gene) %>% summarise(PresentIn = n() ) %>% 
    filter(PresentIn == 6) %>% .$Gene

accuracy.list <- list()
for(i in 6:length(case.index)){
  set.seed(1)
  accuracy.list[[i-5]] <- replicate(10,{
  samples.index <- c( control.index, sample(case.index, i))
  cvsvm(samples.index = samples.index, df_main[,c("Status",feature.genes)], nfolds = 6)
  })
}
tmp <- lapply(accuracy.list, rowMeans) %>% 
  purrr::reduce(.x = ., .f = rbind) %>% as.data.frame(.) %>% 
  dplyr::rename(Accuracy = V1, TrueControls = V2, MCC = V3) %>% 
  mutate(Features = "14-medium", MelSamples = c(6:length(case.index)))

svm.results <- bind_rows(svm.results,tmp)

#Applying a strict threshold to select the feature genes
feature.genes <- genes.mel.healthy %>% filter(Set != "From all") %>% 
  group_by(Set, Gene) %>% summarise(n = n()) %>% 
    filter(n > threshold[3]) %>% 
    ungroup() %>% 
    group_by(Gene) %>% summarise(PresentIn = n() ) %>% 
    filter(PresentIn == 6) %>% .$Gene

accuracy.list <- list()
for(i in 6:length(case.index)){
  set.seed(1)
  accuracy.list[[i-5]] <- replicate(10,{
  samples.index <- c( control.index, sample(case.index, i))
  cvsvm(samples.index = samples.index, df_main[,c("Status",feature.genes)], nfolds = 6)
  })
}
tmp <- lapply(accuracy.list, rowMeans) %>% 
  purrr::reduce(.x = ., .f = rbind) %>% as.data.frame(.) %>% 
  dplyr::rename(Accuracy = V1, TrueControls = V2, MCC = V3) %>% 
  mutate(Features = "2-strict", MelSamples = c(6:length(case.index)))

svm.results <- bind_rows(svm.results,tmp)


write.csv(svm.results, file.path(data.dir,"svm_mel_healthy.csv"),quote = FALSE)
```

```{r SVM Melanoma VS Healthy plots}
#threshold values for genes selection
threshold <- c(260, 540, 800)

#Load the analysis of the chunk above for quick report generation
svm.results <- read.csv(file.path(data.dir,"svm_mel_healthy.csv")) %>% 
  dplyr::select(-1)

ggplot(data = svm.results, aes(MelSamples, Accuracy, color = Features, shape = Features))+
  geom_point(alpha = 0.8)+
  geom_smooth()+
  labs(title = "Mean Accuracy",
       subtitle = "Mean of 6-fold cross validation repeated 10 times",
       x = "Number of melanoma samples to train the model",
       y = "Accuracy (100%)")

ggplot(data = svm.results, aes(MelSamples, MCC, color = Features, shape = Features))+
  geom_point(alpha = 0.8)+
  geom_smooth()+
  labs(title = "Mean MCCs",
       subtitle = "Mean of 6-fold cross validation repeated 10 times",
       x = "Number of melanoma samples to train the model",
       y = "Mathews Correlation Coefficient")

ggplot(data = svm.results, aes(MelSamples, TrueControls, color = Features, shape = Features))+
  geom_point(alpha = 0.8)+
  geom_smooth()+
  labs(title = "Mean True Positives",
       subtitle = "Mean of 6-fold cross validation repeated 10 times",
       x = "Number of melanoma samples to train the model",
       y = "Mean True Positives")

```

#Multi-Classification of the melanoma mutations

**IGNORE THE FOLLOWING. PROBLEM WITH MARKDOWN, NOT EVERYTHING PRESENTED**
**IGNORE THE FOLLOWING. PROBLEM WITH MARKDOWN, NOT EVERYTHING PRESENTED**
**IGNORE THE FOLLOWING. PROBLEM WITH MARKDOWN, NOT EVERYTHING PRESENTED**

~~After finding the genes that can be used as selectors between healthy and cancerous melanoma cells, it is imperative to also try and find a model to classify the different melanoma mutations. Each mutation has a different cancer mechanism, therefore, for a target therapy to work, the different targets of each mutations have to be identified.~~

```{r Functions for LDA}

#Remove collinear
remove.collinear <- function(df, pearson){
  correlation.matrix = cor(df[,-1])
  correlation.matrix[!lower.tri(correlation.matrix)] <- 0
  index <- c(FALSE, apply(correlation.matrix, 2, function(x) any(abs(x)>pearson)))
  return( df[ ,!(index)] )
}

remove.collinear.index <- function(df, pearson){
  correlation.matrix = cor(df[,-1])
  correlation.matrix[!lower.tri(correlation.matrix)] <- 0
  index <- c(FALSE, apply(correlation.matrix, 2, function(x) any(abs(x)>pearson)))
  return(index)
}

#crossV function for K-fold cross-validation
cvlda <- function(df, nfolds){
  folds = createFolds(df$Mutation, k = nfolds)
  accuracy = 0
  class.vector = 0
  for(i in 1:length(folds)){
    train = df[-folds[[i]], ] 
    test = df[folds[[i]], ]
    
    index.train = remove.collinear.index(train, 0.9)
    index.test = remove.collinear.index(test, 0.9)
    index = index.test | index.train
    
    train = train[,!index]
    test = test[,!index]
    
    
    model_lda <- lda(Mutation ~ ., train)
    predict_lda <- predict(model_lda, test[,-1])
    conf.matrix = confusionMatrix(predict_lda$class, factor(test$Mutation))
    accuracy = accuracy + conf.matrix$overal['Accuracy']
    class.vector = class.vector + diag(conf.matrix$table)/table(test$Mutation)
  }
  return(c(accuracy, class.vector)/nfolds)
}


```


```{r LDA mutations feature genes}
library(caret)
library(MASS)

# df_main is the data frame that will be used in all downstream analysis
tmp <- pheno.global %>% 
  column_to_rownames(var = "geo_accession") %>% 
  dplyr::select(Mutation)

df_main <- global.exp.corrected %>% column_to_rownames(var = "SYMBOL") %>% 
  t(.) %>% as.data.frame(.) %>% 
  cbind(tmp, .)# %>% filter(Mutation != "control")
index <- which(df_main$Mutation == "control")
df_main <- df_main[-index,]

#Plotting to show importance of removing collinear
# get_upper_tri <- function(cormat){
#     cormat[lower.tri(cormat)]<- NA
#     return(cormat)
#   }
# df <- get_upper_tri(cor(df_main[,c(feature.genes[[3]])])) %>% 
#   as.data.frame(.) %>% rownames_to_column(var = "Var1") %>% 
#   gather("Var2", "value", -1, na.rm = TRUE)
# 
# ggplot(data = df, aes(Var1, Var2, fill = value))+
#   geom_tile(color = "white")+
#  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
#    midpoint = 0, limit = c(-1,1), space = "Lab", 
#    name="Pearson\nCorrelation") +
#   theme_minimal()+ 
#  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
#     size = 12, hjust = 1))+
#  coord_fixed()


#Feature (gene) selection has to be done.
#A list containing all the different sources for the feature genes has to be created in order
# for the procedure to be better understood.
feature.genes <- list()
#For starters trying with the DEGs when direct contrasting between the mutations w/o wt.
feature.genes$ContrastMutMut <- genes.mut.mut %>% 
  filter(adj.P.Val <0.05 & abs(logFC) > 1 & !grepl("wt", Set)) %>% 
  group_by(Gene) %>% summarize(PresentIn = n()) %>% .$Gene

# Including wt
feature.genes$ContrastMutMutWt <- genes.mut.mut %>% 
  filter(adj.P.Val <0.05 & abs(logFC) > 1 ) %>% 
  group_by(Gene) %>% summarize(PresentIn = n()) %>% .$Gene

# Now trying with  the difference of the DEGs in each mutations
# contrast 
feature.genes$MutHealthy <- genes.mut.healthy %>% 
  group_by(Set, Gene) %>% summarize(n = n() ) %>% 
  filter(n>threshold[3]) %>% ungroup %>% 
  group_by(Gene) %>% summarise(PresentIn = n()) %>% 
  filter(PresentIn == 1) %>% .$Gene

```


```{r LDA mutations, eval = FALSE, include=FALSE}
# The feature genes have been selected
# Trying with all three cases
lda.results <- list()

set.seed(1)
lda.results[[1]] <- replicate(10, {
  samples.index <- c( which(df_main$Mutation != "BRAF"), 
                    sample( which(df_main$Mutation == "BRAF"), 22))     #0.6523
  cvlda( df_main[samples.index,c("Mutation", feature.genes[[1]])], 10)
  })

set.seed(1)
lda.results[[2]] <- replicate(10, {
  samples.index <- c( which(df_main$Mutation != "BRAF"), 
                    sample( which(df_main$Mutation == "BRAF"), 22))     #0.43
  cvlda( df_main[samples.index,c("Mutation", feature.genes[[2]])], 10)
  })

set.seed(1)
lda.results[[3]] <- replicate(10, {
  samples.index <- c( which(df_main$Mutation != "BRAF"), 
                    sample( which(df_main$Mutation == "BRAF"), 22))     #0.40
  cvlda( df_main[samples.index,c("Mutation", feature.genes[[3]])], 10)
  })

names(lda.results) <- names(feature.genes)

lda.results <- lapply( lda.res, function(x) as.data.frame(t(x))) %>% 
  map2_df(.x = ., 
          .y = gsub("Contrast", "", names(feature.genes))
          , ~mutate(.x, Set = .y))

write.csv(lda.results, file.path(data.dir, "lda_results_mutations.csv"), quote = FALSE, row.names = FALSE)
```
  
  
```{r LDA mutations plots}
lda.results <- read.csv(file.path(data.dir, "lda_results_mutations.csv"), stringsAsFactors = FALSE)

#Visualizing the results of the above analysis
df <- lda.results %>% 
  gather(key = MeasuredVariable, value = Measurement, -6)
tmp <- df %>% 
              group_by(Set, MeasuredVariable) %>% 
              summarise(Mean = mean(Measurement) )

ggplot(data = df, aes(MeasuredVariable, Measurement))+
  geom_dotplot( aes(fill = Set, color = Set),
                binaxis = "y", stackdir = "center", alpha = 0.7)+
  stat_summary(fun.y = mean, fun.ymin = mean, fun.ymax = mean,
                 geom = "crossbar",aes(group = Set), 
               alpha = 0.5, width = 0.2, show.legend = FALSE)+
  viridis::scale_fill_viridis(discrete = TRUE)+
  viridis::scale_color_viridis(discrete = TRUE)+
  labs(title = "LDA results",
       x = "Variable of Interest",
       y = "Result",
       subtitle = "10 replications of 10-fold cross validation",
       legend = "Feature Genes",
       caption = "For the BRAF/NRAS/PTEN/wt cases the result shown is the percentage of correctly classified cases") +
  theme(plot.caption=element_text(hjust = -0.01))
```


```{r LDA mutations optimization}
#To see if we can find a better combination of genes as selectors
# we run with the direct contrasts but including one other
# Previous accuracy 0.652

lda.optim <- function(features, extra.features, df){
  accuracy = NULL
  for(i in 1:length(extra.features)){
    new.features = unique(c(features, extra.features[i]))
    set.seed(1)
    accuracy[i] = mean((replicate(5, {
        samples.index = c( which(df$Mutation != "BRAF"), 
                    sample( which(df$Mutation == "BRAF"), 22))     
        cvlda( df[samples.index,c("Mutation", new.features)], 10)
      }))[1,])
  }
  return(c(extra.features[which.max(accuracy)], max(accuracy)))
}

lda.optim.auto <- function( features, extra.features, df){
  v1 = NULL
  v2 = NULL
  optim.features = features
  candidate.features = extra.features
  for( i in 1:20) {
    tmp <- lda.optim(optim.features, candidate.features, df)
    v1[i] <- tmp[1]
    v2[i] <- as.numeric(tmp[2])
    optim.features <- c(optim.features, v1[i])
    candidate.features = candidate.features[which(candidate.features != v1[i])]
  }
  return(data.frame(Gene = v1, Accuracy = v2, stringsAsFactors = FALSE))
}

# system.time(
#   feature.genes.optim <- lda.optim.auto(features = feature.genes[[1]], 
#       extra.features = feature.genes[[2]][-which(feature.genes[[2]] %in% feature.genes[[1]])],
#                                 df = df_main))   #3.4 hours

#write.csv(feature.genes.optim2, file.path(data.dir, "feature_genes_mutations.csv"), quote = FALSE)
tmp <- read.csv(file.path(data.dir, "feature_genes_mutations.csv"),
                                    stringsAsFactors = FALSE)%>% 
  dplyr::select(Gene, Accuracy) %>% 
  bind_rows(data.frame(Gene = "Initial", 
                       Accuracy = mean(lda.results[1:10,1]), stringsAsFactors = FALSE),.)

feature.genes.mutations <- tmp  %>% 
  mutate(Nfeatures = c(4:(dim(tmp)[1]+3))) %>% 
  mutate(Accuracy = as.numeric(format(Accuracy*100, digits = 4)))

ggplot(data = feature.genes.mutations, aes(Nfeatures, Accuracy))+
  geom_point(alpha = 0.7)+
  geom_point(data = feature.genes.mutations[which.max(feature.genes.mutations$Accuracy),],
             aes(Nfeatures, Accuracy), col = "red")+
  ggrepel::geom_label_repel(data = feature.genes.mutations[which.max(feature.genes.mutations$Accuracy),],
             aes(Nfeatures, Accuracy, label = Accuracy))+
  geom_smooth(method = "loess")+
  labs(title = "Accuracy of LDA",
       subtitle = "Based on number of features (genes)")

```
